\vspace{-0.25em} \thetitle

### 2008-01-11 paper.tex
Initial git repository.

\newcommand{\thetitle}{Characterizing Flow Behavior in WLANs}
%\newcommand{\thesubtitle}{Change-Point Detection, Distribution Clustering,\\and Principal Component Analysis}

@@

\title{
\vspace{-0.25em}
\thetitle
%\vspace{0.25em}
%\LARGE{\thesubtitle}
}
\author{
{\large{Stefan~Karpinski, Elizabeth~M.~Belding, Kevin~C.~Almeroth}} \vspace{0.25em}\\
Department of Computer Science \\
University of California, Santa Barbara \vspace{0.35em}\\
\textit{\{sgk,ebelding,almeroth\}@cs.ucsb.edu}
%\vspace{-0.5em}
}

@@

\newcommand{\X}{\mathsf{X}}
\newcommand{\M}{\mathsf{M}}
\newcommand{\E}[1]{\left<#1\right>}
\newcommand{\abs}[1]{\left|#1\right|}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Q}{\mathcal{Q}}
\newcommand{\ceil}[1]{\left\lceil#1\right\rceil}
\newcommand{\floor}[1]{\left\lfloor#1\right\rfloor}

@@

%\begin{abstract}
% We present a technique for representing packet-level flow behavior in a small number of dimensions, while preserving the essential similarities and dissimilarities between flows.
%\end{abstract}

@@

Traffic patterns in wireless local-area networks (\caps{WLAN}s) have recently become recognized as a subject of significant interest and importance in wireless research~\cite{Papadopouli05,Hernandez06:wlan-traffic,Ploumidis07,Karaliopoulos07,Karpinski07:realism,Karpinski07:cbr-failure}. Most of this research has taken a top-down approach, aiming to reproduce high-level statistical characteristics of observed workload patterns. In particular, Her\'andez-Campos~\textit{et~al.}~\cite{Hernandez06:wlan-traffic} showed convincingly that the following statistical models apply to \caps{WLAN} traffic: user arrivals (sessions) in general follow a time-varying Poisson model; the number of flows per session follows a BiPareto distribution; the sizes of flows also follow a BiPareto distribution; the intervals between the initiations of flows within each session follow a Lognormal distribution. These models provide an excellent high-level overview of the behavior of users and applications in \caps{WLAN}s.

The methodology for generating synthetic \caps{WLAN} workload remains incomplete, however. While the high-level scaffolding for producing traffic exists, the low-level behavior of flows is neither understood nor reproducible. The common practice in workload generation is to use a uniform constant bit-rate (\caps{CBR}) model for the packet-level behavior of flows: all flows have the same number of packets, all packets have identical size, and the intervals between packets in each flow are fixed. Karpinski~\textit{et~al.}~\cite{Karpinski07:realism,Karpinski07:cbr-failure} showed, however, that all variants of \caps{CBR} packet behavior models drastically distort important performance metrics, and thus fail the litmus test for realism. Distorted performance resulted even when \caps{CBR} was applied with high-level behavior taken directly from traces \textit{and} using accurate packet count, packet size, and inter-packet interval for each flow. More sophisticated models of packet-level flow behavior are needed before accurate simulation results can be achieved.
%Moreover, in this work, it was assumed that not only how many packets each flow consisted of, but also the total duration of each flow, and thus its average data rate, were both accurately modeled. While this research did not specifically examine the mixed-model case where only one of these parameters was known, this additional ignorance can only worsen the accuracy the model.

Using variable bit-rate (\caps{VBR}) flows is a common elaboration upon the \caps{CBR} flow model. In \caps{VBR} models, the packets sizes and the inter-packet intervals of each flow are separately randomly sampled from independent, identically distributed (i.i.d.) pre-specified distributions.
% independently randomly sampled from fixed, pre-specified distributions. %Both the packet sizes and inter-packet intervals are thus modeled as separate, independent and identically distributed (i.i.d.) time series.
One of the most significant results of~\cite{Karpinski07:realism} is that if high-level workload is realistic, then an i.i.d. \caps{VBR} model for flow behavior accurately reproduces network performance---\textit{so long as realistic packet size and inter-packet interval distributions are used for each flow}. This requirement, however, is non-trivial to satisfy. The size and interval distributions used in~\cite{Karpinski07:realism} were not simple parametric distributions typically used with \caps{VBR} models. Rather, they were empirical distributions taken directly from the observed behavior of each flow in the original trace. Thus, every flow has a unique signature of packet sizes and inter-packet intervals. To produce a realistic \textit{collection} of such signatures, requires an understanding of what realistic signatures ``look like,'' and how frequently they occur relative to each other.

In this paper we use a series of advanced non-parametric statistical techniques
%---including a novel distribution clustering method---
to analyze, understand, characterize, and visualize the space of packets size and inter-packet interval distributions found in real-world examples of \caps{WLAN} traffic. This work complements that of Hern\'andez-Campos~\textit{et~al.}: their session and flow models provide the high-level traffic behavior, while our research explores the low-level details of behavior.% Taken together, these models provide, for the first time, a full-stack composite model capable of generating statistically realistic \caps{WLAN} workload.

\todo{summary of specific findings}

\todo{paper outline paragraph}

@@

%Previous work has shown that with knowledge of the following properties for each flow, realistic performance of wireless traffic can be reproduced accurately:
%\begin{enumerate}
%\item packet count (equivalently, duration),
%\item distribution of packet sizes,
%\item distribution of inter-packet intervals.
%\end{enumerate}
%In short, for the purpose of generating experimental workload that accurately predicts performance under real-world traffic conditions, it is sufficient to know the marginal distributions of packet properties, without any time-series analysis. Each simulated flow can be produced by generating a sequence of packets in the following manner: randomly generate a packet size by sampling the size distribution; wait a randomly generated duration sampled from the inter-packet interval distribution; repeat until the appropriate number of packets has been generated.

\section{Methodology}\label{sec:methodology}

Our goal in this research is to provide an effective general methodology for analyzing, understanding, and characterizing the subspace of packet size and inter-packet interval distributions found across the flows in \caps{WLAN} trace traffic. In essence, we require a ``distribution of distributions.'' %Essentially, one needs to find the subspace of all possible signatures in which realistic signatures occur, and how often.
However, size and interval distributions exist in spaces with very high dimensionality. In theory, both types of distributions live in infinite-dimensional Hilbert spaces.%TODO: verify this
\footnote{The space of size distributions has a countably infinite number of dimensions, while the space of inter-packet interval distributions has an uncountably infinite number of dimensions.}
In practice, limits on possible values and quantization make the effective dimensions finite, but nonetheless too large for simple, direct analysis.% of the regions in which realistic flow behaviors can be found.

To reduce the dimensionality of distribution data, we use a sequence of techniques.
First, we divide each flow's series of values---sizes or inter-packet intervals---into segments of homogeneous behavior, using time-series change-point detection. We call these subsegments of flows ``\textit{flowlets}.'' Second, we apply agglomerative clustering to find closely related behaviors. Third, we use principal component analysis (\caps{PCA}) to isolate the most important aspects of behavior across the entire collection. The resulting transformation allows us to meaningfully visualize the space of flow behaviors with respect to packet size or inter-packet intervals in one, two or three dimensions. With each additional dimension, more information about the behavior is captured. %Finally, we analyze the geometry and distribution of behaviors with respect to their principal components.

Using clustering followed by \caps{PCA} is a common approach to analyzing data with high-dimensionality. % TODO: cite this
The nature of our data, however, requires novel techniques---these are not standard multidimensional data-points for which normal clustering techniques or standard \caps{PCA} can be utilized. Rather, the data are variably sized samples from unknown, non-parametric distributions. Our clustering and \caps{PCA} methods are specially designed, based on prior art in non-parametric statistics, to address the difficulties encountered in this problem.

\subsection{The Baumgartner-Wei\ss-Schindler Test}

@@

In the case of packet sizes and inter-packet intervals, there are no ``clean'' parametric models that capture the variety of behaviors seen in network traces. Therefore, we rely on non-parametric statistical procedures for our analysis. There are a number of non-parametric two-sample goodness-of-fit tests in the statistical literature. Such tests take two samples of scalar values and test the null hypothesis that the samples were drawn from the same (unknown) underlying distribution. Non-parametric tests do not make any assumptions about the shape or nature of the underlying distributions, other than that they are continuous and real-valued.

We use the Baumgartner-Wei\ss-Schindler (\caps{BWS}) test~\cite{BWS98} rather than the more well-known Kolmogorov-Smirnov (\caps{KS}) or Cram\'er-von Mises (\caps{CvM}) tests for three reasons.\footnote{There is a generalized version of the CvM test~\cite{Pettitt79} that may compare more favorably with the BWS test, but no simulated results were presented by which to evaluate its power. This generalized CvM and BWS are closely related: both use the same variance-weighted CvM integral as their basis, but they use different estimators of this quantity as test statistics.} First, the \caps{BWS} test is experimentally more powerful for detecting differences in both location and shape, performing very nearly as well as parametric tests, in cases where such specialized tests exist. Second, in certain cases both the \caps{KS} and \caps{CvM} tests suffer from limited power as sample size grows: even when the sample size is allowed to go to infinity, for some types of differences, the test power levels off well before reaching unity. This causes many failures to reject the null hypothesis even when samples are large and significant underlying differences exist. The \caps{BWS} test does not exhibit this effect. Finally, the exact distribution of the \caps{BWS} statistic approaches its asymptotic distribution very quickly: the asymptotic distribution is already a good approximation for samples as small as ten values. This is highly desirable since the cost of computing exact distributions grows exponentially with sample size, so the asymptotic distribution must be used even for modestly sized samples.

@@

\subsection{Flow Splitting}

The first analytical task is to split the packet size and inter-packet interval time series for each flow into segments with homogeneous statistical behavior. A great deal of classical statistical research has been done in the general area of time-series change-point detection~\cite{Basseville93}. These classical approaches, however, cannot be applied when the number of change points are unknown and, worse still, the underlying models for the stochastic processes are unknown. Work has been done in econometrics that allows a single change point to be discovered using non-parametric \caps{KS}- or \caps{CvM}-type significance tests~\cite{Inoue01}. Zheng~\textit{et~al.}~\cite{Zhang01} use recursive iteration of this procedure with the Fligner-Policello (\caps{FP}) test to detect change points in Internet path properties.
Our technique is the same except that we use the \caps{BWS} test rather than \caps{FP}.\footnote{Strictly, our ``backward elimination'' post-processing procedure differs: we eliminate the insignificant change point with the largest $p$-value first and iterate; Zhang~\textit{et~al.} eliminate change points in the order they are encountered.}
The \caps{FP} test detects only changes in central location (non-parametrically), whereas \caps{BWS} detects changes in both location and shape.

\subsection{Flowlet Clustering}

\subsection{Principal Component Analysis}

\section{Results}\label{sec:results}

\section{Conclusions}\label{sec:conclusions}

\section{Acknowledgments}
This work was funded in part by NSF Career Award CNS-0347886 and by NSF NeTS Award CNS-0435527. Special thanks to Matthew Allen, Khaled Harras, Allan Knight, and Rich Wolski for their invaluable help and advice.

\bibliography{IEEE,references}

>>>>

\newcommand{\thetitle}{An Atomic Theory of Flow Behavior}

@@

\title{
\vspace{-0.25em}
\thetitle
}
\author{
{\large{Stefan~Karpinski, Elizabeth~M.~Belding, Kevin~C.~Almeroth}} \vspace{0.25em}\\
Department of Computer Science \\
University of California, Santa Barbara \vspace{0.35em}\\
\textit{\{sgk,ebelding,almeroth\}@cs.ucsb.edu}
%\vspace{-0.5em}
}

@@

\newcommand{\E}[1]{\left<#1\right>}
\newcommand{\abs}[1]{\left|#1\right|}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Z}{\mathbb{Z}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\ceil}[1]{\left\lceil#1\right\rceil}
\newcommand{\floor}[1]{\left\lfloor#1\right\rfloor}

@@

\begin{abstract}
We propose an entirely new approach to understanding and analyzing the behavior of flows in packet networks. The essential concept of this new approach is to find atomic units of time and size behavior in traces of network flows. While the nonparametric statistical techniques for extracting these basic behavioral units are complex, the end result is quite simple: the units provide an alphabet for flow behavior. From a finite set of behavioral units, an infinite variety of actual behaviors can be composed. The space of behaviors naturally becomes a vector space generated by these atomic units of behavior. We use numerical linear algebra techniques to demonstrate useful and immediate applications of this theory to real-time traffic analysis, anomaly and attack detection, as well as workload generation for wireless experiments.
\end{abstract}

@@

The trouble with trying to understand or model behavioral patterns in packet networks is that beyond the packets themselves there is no inherent behavioral structure. There are flows of packets with the same source and destination \caps{IP} addresses and \caps{TCP/UDP} port numbers, and sessions of flows belonging to the same host, but these imply only very limited behavioral similarity. Each one has its own unique sequence of packet sizes and inter-packet intervals with no obvious relation to each other. Without fundamentally behavioral elements of structure, traffic traces are just ``packet soup,'' so to speak. Accordingly, traditional approaches to traffic analysis have either focused on aggregate traffic measures or categorized flows by well-known port numbers and application types. Some fiendishly clever techniques have been proposed to tease out the applications types underlying traffic found in network traces~\cite{Karagiannis05}. Without an inherently behavioral theory of network traffic, however, we believe that insight into the fine structure of network behavior is ultimately very limited.

%Such indirect approaches, however, can ultimately provide only limited insight into the fine structure of network behavior.

We propose to turn this problem on its head by providing an atomic behavioral theory for network traffic. We begin with the concept of a packet flow as a natural starting point and define a ``{flowlet}'' as a segment of a flow with statistically consistent packet size or inter-packet interval behavior. %\footnote{The word ``or'' is not a typo: there are two independent types of flowets.}
Common flowlet behaviors are extracted from a body of traffic traces, using statistical clustering. These clusters of similarly behaved flowlets provide the atomic units of flow behavior: by mixing basic behaviors from a finite collection of flowlet clusters in varying proportions, an infinite variety of flow behaviors emerge. It becomes natural to view the space of flow behaviors as a vector space of linear combinations of flowlet clusters.

To demonstrate the viability and utility of this atomic theory of flow behavior, we apply standard numerical linear algebra techniques to the resulting vector space of flow behaviors. Principal component analysis (\caps{PCA}) allows us to reduce the dimensionality of the vector space of flow behaviors while preserving the vast majority (99\%) of the variatibility in the original data. \caps{PCA} allows us to represent flow behaviors concisely in only eight dimensions ($\R^8$). Moreover, this representation has several desirable properties. The eight coordinates of flows mapped into this space are linearly uncorrelated (non-linear dependencies, however, remain). The first dimension of the transformed data captures the most important differences between flows; the second dimensions the next most important, and so on. The first two or three dimensions, thus, are ideal for visualization of behavioral differences. Finally, the dimensions are naturally scaled so that the standard Euclidean distance provides a good measure of the behavioral (dis)similarity between flows. As a result, standard multi-dimensional analysis and modeling techniques, such as $k$-means clustering, can be applied directly to the transformed flow behaviors.

We present two useful, immediate applications of this new theory of flow behavior. First, once a body of traffic has been analyzed and \caps{PCA}-transformation matrices computed, new flows can be mapped into the flow-behavior space using only fast matrix and vector computations with pre-computed matrices. This allows the possibility of completely real-time traffic analysis and visualization.
%
Second, since the coordinates of the \caps{PCA}-transformed flow behaviors are uncorrelated and exhibit limited non-linear dependencies, we can roughly model them using a multivariate normal distribution. Random deviates from this distribution can be mapped back into actual flow behaviors, allowing the generation of an entire network's worth of heterogeneous synthetic flow behaviors with the realistic intra-flow and inter-flow behavior matching the original trace traffic. This ability is of particular importance for experimental wireless research, where it has been shown that using standard, na\"ive traffic models, such as the uniform constant bit-rate (\caps{CBR}) flows, severely distorts important performance metrics at all levels of the network~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.

%Three final characteristics make our analytical methodology particularly promising. First, the techniques and results are described are not dependent on the data we have used: they can be applied to any collection of network traffic.

%, with the dimensions ordered to capture diminishing variance in behavior. Thus, the first two dimensions can be used to provide a faithful visual map of flow behaviors. Standard clustering techniques can be applied to the transformed data to discover behavioral similarities across protocol boundaries. Moreover, once the transformation has been established from real data, it can be quickly applied to new flows using simple matrix operations. Our methodology thereby allows for real-time analysis of network traffic.

%%Common flowlet behaviors are found across all the flows in network traces, thereby providing the basis for an atomic theory of flow behavior.

%%We provide the statistical and computational techniques needed to extract flowlets from traffic traces, we provide the basic building blocks for understanding flow behavior.

%%What we propose is a view of network traffic that let's us directly characterize the \textit{behavior} of flows, and then view the traditional traffic types as annotations to that fundamental characterization.

%% How can one even begin to analyze flow behavior data, when each flow has a completely unique signature of packet sizes and inter-packet intervals. The first part of our analysis methodology addresses this problem by extracting ``atoms'' of flow behavior. First we split each flow into segments of homogeneous behavior, and then we cluster these segments across all the flows to find elements of similar behavior from which all flows are composed. You can think of these clusters an alphabet for flow behavior: flows are words composed by stringing together sequences of letters.

%Traffic patterns in wireless local-area networks (\caps{WLAN}s) have recently become recognized as a subject of significant interest and importance in wireless research~\cite{Papadopouli05,Hernandez06:wlan-traffic,Ploumidis07,Karaliopoulos07,Karpinski07:realism,Karpinski07:cbr-failure}. It has been shown that using common, na\"ive traffic models in experiments can severely distort important performance metrics at all levels of the network, sometimes completely inverting relative performance of protocols~\cite{Karpinski07:cbr-failure}. Significant progress has been made in understanding and modeling the high-level properties of traffic patterns in \caps{WLAN}s~\cite{Papadopouli05,Hernandez06:wlan-traffic,Ploumidis07,Karaliopoulos07}. Finding complete multi-level traffic models that can accurately reproduce the performance characteristics of real-world traffic, however, remains an unsolved problem.

%% Understanding and modeling the complete pattern of traffic in any local-area network is an extremely challenging analytical modeling problem. Transmissions in the network are logically organized into flows, consisting of series of packets between a pair of nodes on a fixed pair of \caps{TCP} or \caps{UDP} port numbers. Thus, each flow is already a related pair of time series, whose behavior can exhibit complex feedback. The network is comprised of a many nodes with heterogeneous behavioral characteristics, and with a vast variety of flows connecting them. The flows themselves are described by distributions of packet sizes and inter-packet intervals, but across the entire collection of flows and nodes, one finds distributions of these distributions. In short, the ``curse of dimensionality'' which plagues so many modeling problems appears in spades.

%%\todo{discuss challenges and applications}

%In this paper we use a series of advanced statistical techniques to analyze, understand, and visualize the space of packet size and inter-packet interval distributions found in real-world \caps{WLAN} traffic. Our work complements the top-down approach that most research in this area has taken. In particular, the work of Hern\'andez-Campos~\textit{et~al.} provides high-level session and flow models that give a convincing overview of network traffic. Karpinski~\textit{et~al.}~\cite{Karpinski07:realism} showed that if, in addition to such high-level models, the distributions of packet sizes and inter-packet intervals across all flows can be realistically modeled, then whole-network performance characteristics can be accurately reproduced in simulation.

%Reproducing a network's worth of heterogeneous flow behaviors in a manner that resembles reality is easier said than done, however. The multi-level complexity, non-linear feedback, and the very high dimensionality of the spaces involved conspire to make the analysis seem all but intractable. In the last decades, however, a great deal of work has been done in fields with similar difficulties: biometrics, econometrics, genetics, ecology, meteorology. The common tools of these fields, when traditional methods fail, are nonparametric statistics, clustering, and a variety of dimensionality reduction techniques. We employ all of these to extract order from the chaos of network flow behavior.

%Our processing methodology transforms raw trace data through a series of stages such that common structure is extracted at every stage, while each flow's individual behavior can still be reconstructed without significant loss of accuracy. %\footnote{Our threshold for loss of accuracy at each transformation is 1\%. We use a standard significance level of $\alpha=0.05$ for our statistical tests.}
%In overview, the procedure first uses time series change point detection to extract basic elements of time and size behavior from raw flow data. Each flow is split into ``flowlets''---sequences of packets with homogeneous statistical behavior. The flowlets are then clustered across flows, extracting common elements of behavior for the entire data set. Finally, we apply standard numerical linear algebra techniques to further reduce the dimensionality of the representation, until each flow can  be expressed with only eight real-valued coefficients.

%The transformed representation of flow behaviors has many useful properties. The coordinates of the transformed data are uncorrelated and normalized, with the dimensions ordered to capture diminishing variance in behavior. Thus, the first two dimensions can be used to provide a faithful visual map of flow behaviors. Standard clustering techniques can be applied to the transformed data to discover behavioral similarities across protocol boundaries. Moreover, once the transformation has been established from real data, it can be quickly applied to new flows using simple matrix operations. Our methodology thereby allows for real-time analysis of network traffic.

%Finally, since the coefficients of the transformed flows are guaranteed to be uncorrelated with zero mean---and because they happen to exhibit limited dependence---we can roughly approximate their distribution as a standard multivariate normal. Random deviates of this approximating distribution can be transformed back into flow behaviors, allowing us to randomly generate collections of flows in a more realistic manner than has previously been achieved.

%%The rest of the paper is organized as follows. In Section~\ref{sec:motivation} we present motivation and related work. Our analytical methodology is presented in Section~\ref{sec:methodology}, while the results of our experiments are explained and analyzed in Section~\ref{sec:results}. Finally, in Section~\ref{sec:conclusions}, we conclude with the impact of and future directions for this research.

%%This work is a collection of advanced analytical techniques for exploring and understanding the fine structure of flow behavior across a large body of traffic. The specific results are of less importance than the techniques themselves, which are far from obvious, and can be applied to many different data sets---and even in other domains.

\section{Motivation \& Related Work}
\label{sec:motivation}
\label{sec:related-work}

In Internet traffic analysis, the detailed structure of workload patterns in local-area networks (\caps{LAN}s) is of limited interest. 
%Bottlenecks in the Internet are found in the core of the Internet or at the last mile.
Capacity has become so cheap and plentiful in wired \caps{LAN}s that workload details are simply irrelevant in the face massive over-provisioning. Wireless networks, however, are fundamentally different: the entire medium is at the ``edge'' of the network and the most basic resources of bandwidth and power are severely limited, with no permanent relief in sight. Thus, over-provisioning is simply not an option, and the fine-grained details of traffic patterns have been shown to have a dramatic impact on network performance~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
As a result, modeling and analysis of wireless \caps{LAN} (\caps{WLAN}) traffic has recently become a hot topic in the wireless community, and a significant body of high-quality research on this subject has been produced~\cite{Papadopouli05,Hernandez06:wlan-traffic,Ploumidis07,Karaliopoulos07,Karpinski07:realism,Karpinski07:cbr-failure}.

Most of the recent surge of research in \caps{WLAN} traffic has taken a top-down approach, using parametric models to reproduce high-level statistical characteristics of observed workload behavior~\cite{Papadopouli05,Hernandez06:wlan-traffic,Ploumidis07,Karaliopoulos07}. In particular, Her\'andez-Campos~\textit{et~al.}~\cite{Hernandez06:wlan-traffic} showed convincingly that the following models are applicable to \caps{WLAN} traffic: user arrivals (sessions) follow a time-varying Poisson model; the number of flows per session follows a BiPareto distribution; the sizes of flows follow a BiPareto distribution; the intervals between the initiations of flows within each session follow a Lognormal distribution. These models provide an excellent and convincing high-level overview of the behavior of users and applications in \caps{WLAN}s.

The methodology for generating synthetic workload remains incomplete, however. While the high-level scaffolding for producing \caps{WLAN} traffic exists, the low-level behavior of flows is neither understood nor reproducible. The common practice in workload generation is to use a uniform constant bit-rate (\caps{CBR}) model for the packet-level behavior of flows: all flows have the same number of packets, all packets have identical size, and the intervals between packets in each flow are of a single, fixed duration. % TODO: citation describing CBR
Karpinski~\textit{et~al.}~\cite{Karpinski07:cbr-failure}, however, showed that all types of \caps{CBR} packet behavior models drastically distort important performance metrics, and thus fail the litmus test for realism. Distorted performance resulted even when \caps{CBR} was applied with high-level behavior taken directly from traces \textit{and} using accurate packet count, average packet size, and average inter-packet interval for each flow~\cite{Karpinski07:realism}.% More sophisticated models of packet-level flow behavior are needed before accurate simulation results can be achieved.
%Moreover, in this work, it was assumed that not only how many packets each flow consisted of, but also the total duration of each flow, and thus its average data rate, were both accurately modeled. While this research did not specifically examine the mixed-model case where only one of these parameters was known, this additional ignorance can only worsen the accuracy the model.

Using variable bit-rate (\caps{VBR}) flows is a common elaboration upon the \caps{CBR} flow model. In \caps{VBR} models, the packets sizes and the inter-packet intervals are each randomly sampled from pre-specified, independent, identically distributed (i.i.d.) distributions. % TODO: citation for VBR
% independently randomly sampled from fixed, pre-specified distributions. %Both the packet sizes and inter-packet intervals are thus modeled as separate, independent and identically distributed (i.i.d.) time series.
One of the most significant results of~\cite{Karpinski07:realism} is that if high-level workload is realistic, then an i.i.d. \caps{VBR} model for flow behavior accurately reproduces network performance---with one \textit{very} major caveat: each flow must have its own unique, realistic signature of packet size and inter-packet interval distributions. Producing a realistic cross-network collection of these distributions is, to date, an unsolved problem. The size and interval distributions used in~\cite{Karpinski07:realism} were empirical estimates, taken directly from the observed behavior of each flow in the original trace. This result has a very important implication for the present work: \textit{to characterize flow behavior, it is sufficient to know each flow's distribution of packet sizes and inter-packet intervals}. It is precisely this problem of understanding, modeling, and recreating realistic collections of packet size and inter-packet interval distributions for an entire network that this paper addresses.

% This requirement, however, is non-trivial to satisfy. The size and interval distributions used in~\cite{Karpinski07:realism} were not simple parametric distributions typically used with \caps{VBR} models. Rather, they were empirical distributions taken directly from the observed behavior of each flow in the original trace. Thus, every flow has a unique signature of packet sizes and inter-packet intervals. To produce a realistic \textit{collection} of such signatures, requires an understanding of what realistic signatures ``look like,'' and how frequently they occur relative to each other.

@@

%Previous work has shown that with knowledge of the following properties for each flow, realistic performance of wireless traffic can be reproduced accurately:
%\begin{enumerate}
%\item packet count (equivalently, duration),
%\item distribution of packet sizes,
%\item distribution of inter-packet intervals.
%\end{enumerate}
%In short, for the purpose of generating experimental workload that accurately predicts performance under real-world traffic conditions, it is sufficient to know the marginal distributions of packet properties, without any time series analysis. Each simulated flow can be produced by generating a sequence of packets in the following manner: randomly generate a packet size by sampling the size distribution; wait a randomly generated duration sampled from the inter-packet interval distribution; repeat until the appropriate number of packets has been generated.

\section{Methodology}
\label{sec:methodology}

Our goal is to provide an effective general methodology for analyzing and characterizing the space of packet size and inter-packet interval distributions found across the flows in \caps{WLAN} trace traffic. In essence, we require a ``distribution of distributions.'' %Essentially, one needs to find the subspace of all possible signatures in which realistic signatures occur, and how often.
However, size and interval distributions exist in spaces with very high dimensionality. In theory, both types of distributions live in infinite-dimensional Hilbert spaces. %TODO: verify this
%\footnote{The space of size distributions has a countably infinite number of dimensions, while the space of inter-packet interval distributions has an uncountably infinite number of dimensions.}
In practice, limits on possible values and quantization make the effective dimensions finite, but nonetheless too large for simple, direct analysis.% of the regions in which realistic flow behaviors can be found.

%\todo{paragraph about data}

To reduce the dimensionality of distribution data, we use a series of techniques.
First, we divide each flow's sequence of values---sizes or inter-packet intervals---into segments of consistent behavior, using time series change point detection. We call these segments of flows ``flowlets.'' Second, we apply agglomerative clustering to find clusters of flowlets with mutually consistent behaviors. These flowlet clusters are the atomic units that allow us to systematically apply standard techniques from other fields to analyze and understand flow behaviors. Each flow's behavior can be expressed as some weighted sum of constituent flowlet clusters. Thus, the space of flow behaviors have naturally become a vector space generated by the flowlet clusters. In the final stage of processing we use principal component analysis (\caps{PCA}) to reduce dimensionality of this vector space and isolate the most important aspects of behavior across the entire collection.

%The resulting transformation allows us to meaningfully represent the space of flow behavior in only a small number of dimensions (in our case, eight). Even more importantly, the transformation captures the vast majority (99\%) of the variations in flow behavior. Finally, the transformed data can be roughly approximated by a multi-variate normal distribution, and random deviates of this distribution can be mapped back into the original space of flow behaviors. This provides a way to generate collections of random flow behaviors such that the relationships and patterns found in the trace are respected and preserved.
% with respect to packet size or inter-packet intervals in one, two or three dimensions. With each additional dimension, more information about the behavior is captured. %Finally, we analyze the geometry and distribution of behaviors with respect to their principal components.

%Using clustering followed by \caps{PCA} is a common approach to analyzing data with high-dimensionality. % TODO: cite this
%The nature of our data, however, requires novel techniques---these are not standard multidimensional data-points for which normal clustering techniques or standard \caps{PCA} can be utilized. Rather, the data are variably sized samples from unknown, nonparametric distributions. Our clustering and \caps{PCA} methods are specially designed, based on prior art in nonparametric statistics, to address the difficulties encountered in this problem.

\subsection{Nonparametric Goodness-of-Fit Tests}

@@

In the case of packet sizes and inter-packet intervals, there are no ``clean'' parametric models that capture the variety of behaviors seen in network traces. Therefore, we rely on nonparametric statistical procedures for our analysis. There are a number of nonparametric two-sample goodness-of-fit tests in the statistical literature. % TODO: citation here
Such tests take two samples of scalar values and test the null hypothesis that the samples were drawn from the same (unknown) underlying distribution. Nonparametric tests do not make any assumptions about the shape or nature of the underlying distributions, other than that they are continuous and real-valued. Such tests can detect various differences in both location and shape of distributions. We use the Baumgartner-Wei\ss-Schindler (\caps{BWS}) test~\cite{Baumgartner98} rather than the more well-known Kolmogorov-Smirnov (\caps{KS}) or Cram\'er-von Mises (\caps{CvM}) tests because: \caps{BWS} exhibits significantly superior power in simulation; the exact distribution of the \caps{BWS} test statistic for small sample sizes is already very close to its asymptotic distribution, which is untrue for \caps{KS} and \caps{CvM}.

% First, the \caps{BWS} test is experimentally more powerful for detecting differences in both location and shape, performing very nearly as well as parametric tests, in cases where such specialized tests exist.\footnote{There is a generalized CvM-type two-sample test~\cite{Pettitt79} that may compare more favorably with the BWS test. However, no simulated results were presented by which to evaluate its power. This generalized CvM and BWS are closely related: both use the same variance-weighted CvM integral as their basis, but apply different estimators of this integral as test statistics.} Second, in certain cases both the \caps{KS} and \caps{CvM} tests suffer from limited power as sample size grows: even when the sample size is allowed to go to infinity, the test power can level off well before reaching unity. This causes many failures to reject the null hypothesis even when samples are large and significant underlying differences exist. The \caps{BWS} test does not exhibit this effect. Finally, the exact distribution of the \caps{BWS} statistic approaches its asymptotic distribution very quickly: the asymptotic distribution is already a good approximation for samples as small as ten values. This is highly desirable since the cost of computing exact distributions grows exponentially with sample size, so the asymptotic distribution must be used even for modestly sized samples.

@@

\subsection{Flow Splitting}\label{sec:flow-splitting}

The first analytical task is to split the packet size and inter-packet interval time series for each flow into segments with consistent statistical behavior. A great deal of classical statistical research has been done in the general area of time series change point detection~\cite{Basseville93}. These classical approaches, however, cannot be applied when the number of change points are unknown and, worse still, the underlying models for the stochastic processes are unknown. Work has been done in econometrics that allows a single change point to be discovered using nonparametric \caps{KS} or \caps{CvM} type significance tests~\cite{Inoue01}. Zheng~\textit{et~al.}~\cite{Zhang01} use recursive iteration of this procedure with the Fligner-Policello (\caps{FP}) test to detect change points in Internet path properties. Our technique is the same except that we use the \caps{BWS} test rather than \caps{FP}; the \caps{FP} test detects only changes in central location, whereas \caps{BWS} detects changes in both location and shape.%
\footnote{Our ``backward elimination'' post-processing procedure differs slightly too: we eliminate the least significant change point candidate first and iterate until only significant ones remain; Zhang~\textit{et~al.} eliminate insignificant candidates in sequential order. In cases where these procedures produce different results, our algorithm yields more homogeneous behavior between change points.} %We also stop our recursion if either the left or right sets of values becomes smaller than ten. At this point statistical tests lose power, and can easily suffer from both false rejections of the null and false failures to reject the null.
Allen~\textit{et~al.}~\cite{Allen07} use this procedure with the \caps{KS} test to detect change points in time series of network bandwidth measurements.
Space unfortunately does not permit us to reproduce the procedure here.
%We briefly describe the procedure here.
%
%\subsubsection{Candidate Generation}
%For each sequence index, $k$, compute the $p$-value comparing the set of values to the left of $k$ compared with the set of values to the right of $k$. Split the sequence at the value of $k$ with the most significant (smallest) $p$-value and recurse. Terminate the recursion if either half is empty or if none of the $p$-values are significant. Each splitting point is a candidate change point for the second phase.
%
%\subsubsection{Candidate Elimination}
%Iterate through set of candidate change points, computing the $p$-value of the samples of values immediately before and after each candidate (i.e. with no intervening candidates). Of the candidates that are insignificant, eliminate the one with the least significant (largest) $p$-value and iterate. Terminate the iteration when all $p$-values are significant. The remaining candidates are the final change points found by the algorithm.
%
The output of change point detection algorithm is a collection of ``flowlets''---segments of flows where the packet size or inter-packet interval behaviors are statistically consistent. %There may be some mixing of behaviors within each flowlet, but it is non-temporal in the sense that it cannot be separated out by dividing the sequence of values with respect to time.
The flowlet splitting procedure is applied to intervals and sizes independently, so there are two separate collections of flowlets. Splitting and clustering of intervals and sizes are treated separately because~\cite{Karpinski07:realism} indicates that these can be modeled independently without loss of realism.

\subsection{Flowlet Clustering}
\label{sec:flowlet-clustering}

The second stage of our methodology takes the time and size flowlets produced by the first stage and clusters them into groups exhibiting similar behaviors. Again, there is an extensive body of work on classical clustering algorithms~\cite{Jain99}. As before, however, the classical techniques cannot be applied because the data are fundamentally different from those found in classical problems. Our data are not vectors in a classical Euclidean space, but rather random samples from unknown, nonparametric distributions. Two-sample goodness-of-fit tests provide the ability to compare a pair of samples, but to the best of our knowledge there exists no prior art in statistics literature extending such tests to clustering collections of samples into similar groups. A technical note from management science~\cite{Ruefli00}, describes an interative clustering technique based on the \caps{KS} test. Our procedure resembles the agglomerative version of their procedure. We utilize the \caps{BWS} test rather than \caps{KS}, however, for the same reasons as discussed in Section~\ref{sec:flow-splitting}. Our algorithm can be applied with any general two-sample nonparametric goodness-of-fit test, however, since the test is used only as a black-box generator of $p$-values.
